Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach

### 2024-09-26 paper/paper.qmd
Touch Ups and changing readme file

ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2) +
  theme_minimal() +
  labs(x = "Date of Sample Collection", y = "Service User Count") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

@@

Moreover, @tbl-summary-statistics provides key summary statistics on critical variables such as occupancy rate (beds), service user count, and capacity (actual beds). On average, the occupancy rate for beds is quite high at 96.18%, with a standard deviation of approximately 9.35%, indicating that most shelters are consistently near full capacity. Similarly, the mean service user count stands at 38.41, with a high standard deviation of 35.41, suggesting that there is substantial variability in the number of individuals using the shelters across the dataset. Notably, the data for capacity (actual beds) is missing, which may indicate incomplete reporting or issues in data collection regarding shelter bed availability. Further analysis is required to understand the implications of this missing data. The 'Families' sector was excluded from the final visualization due to a significant portion of missing data in the occupancy rate, which resulted in unreliable service user counts. Including this category would have distorted the interpretation of trends across sectors. Therefore, it was deemed appropriate to focus on the other sectors that have more complete and consistent data.

@@

ggplot(yearly_data, aes(x = factor(year), y = Avg_Occupancy_Rate_Beds)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
  labs(x = "year", y = "Average Occupancy Rate (Beds)") +
  theme_minimal() +
  coord_cartesian(ylim = c(80, NA))

@@

ggplot(yearly_bed_capacity, aes(x = factor(year), y = Total_Bed_Capacity, fill = program_model)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "year", y = "Actual Bed Capacity") +
  theme_minimal()

@@

ggplot(monthly_data, aes(x = month, y = service_user_count, color = factor(year), group = year)) +
  geom_line(size = 1) +
  labs(x = "month", y = "Service User Count", color = "Year") +
  theme_minimal() +
  scale_x_discrete(limits = month.abb)
```

@@

```{r, warning=FALSE}
#| label: fig-service-sector
#| fig-cap: Service User Count by sector for Each year
#| echo: false
sector_data <- data %>%
  group_by(year, sector) %>%
  summarise(Total_Service_User_Count = sum(service_user_count, na.rm = TRUE), .groups = "drop")

@@

ggplot(sector_data_filtered, aes(x = sector, y = Total_Service_User_Count, fill = sector)) +
  geom_bar(stat = "identity") +
  labs(x = "sector", y = "Total Service User Count") +
  facet_wrap(~ year, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, NA))

>>>>

ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2, size = 0.1) + # Adjust size here
  theme_minimal() +
  labs(x = "Date of Sample Collection", y = "Service User Count") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

@@

Moreover, @tbl-summary-statistics provides key summary statistics on critical variables such as occupancy rate (beds), service user count, and capacity (actual beds). On average, the occupancy rate for beds is quite high at 96.18%, with a standard deviation of approximately 9.35%, indicating that most shelters are consistently near full capacity. Similarly, the mean service user count stands at 38.41, with a high standard deviation of 35.41, suggesting that there is substantial variability in the number of individuals using the shelters across the dataset. The 'Families' sector was excluded from the final visualization due to a significant portion of missing data in the occupancy rate, which resulted in unreliable service user counts. Including this category would have distorted the interpretation of trends across sectors. Therefore, it was deemed appropriate to focus on the other sectors that have more complete and consistent data.

@@

ggplot(yearly_data, aes(x = factor(year), y = Avg_Occupancy_Rate_Beds)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
  labs(x = "Year", y = "Average Occupancy Rate (Beds)") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

@@

ggplot(yearly_bed_capacity, aes(x = factor(year), y = Total_Bed_Capacity, fill = program_model)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Actual Bed Capacity") +
  theme_minimal()

@@

ggplot(monthly_data, aes(x = month, y = service_user_count, color = factor(year), group = year)) +
  geom_line(size = 1) +
  labs(x = "Month", y = "Service User Count", color = "Year") +
  theme_minimal() +
  scale_x_discrete(limits = month.abb)
```

@@

```{r, warning=FALSE}
#| label: fig-service-sector
#| fig-cap: Service User Count by sector for Each year
#| echo: false
library(scales)

sector_data <- data %>%
  group_by(year, sector) %>%
  summarise(Total_Service_User_Count = sum(service_user_count, na.rm = TRUE), .groups = "drop")

@@

ggplot(sector_data_filtered, aes(x = sector, y = Total_Service_User_Count, fill = sector)) +
  geom_bar(stat = "identity") +
  labs(x = "Sector", y = "Total Service User Count") +
  facet_wrap(~ year, ncol = 2) +
  scale_y_continuous(labels = comma) +  # Format y-axis with whole numbers
  scale_fill_discrete(name = "Sector") +  # Change legend title
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, NA))
